# Data

### Technical description

The only dataset used for this project is the Youth Tobacco Survey (YTS) Data, provided by the Office on Smoking and Health within the National Center for Chronic Disease Prevention and Health Promotion at the Centers for Disease Control and Prevention (CDC). The dataset can be directly accessed online at https://data.cdc.gov/Survey-Data/Youth-Tobacco-Survey-YTS-Data/4juz-x2tp.

The YTS Data is derived from the web-based National Youth Tobacco Survey (NYTS), administered annually by CDC. The survey covers various aspects, including tobacco use, exposure to environmental tobacco smoke, smoking cessation, school curriculum, minors' access to tobacco products, knowledge and attitudes about tobacco, and exposure to pro-tobacco and anti-tobacco media messages. A two-stage cluster sample design was employed to produce representative samples of students in middle schools (grades 6--8) and high schools (grades 9--12). The first stage involves selecting schools with probability proportional to their enrollment size, while the second stage randomly selects classes from the selected schools. All students in the chosen classes are eligible to participate in the survey, and a minimum overall response rate of 60% is required. NYTS has been conducted since 1999, and the YTS Data synthesizes the results from 1999 to 2017.

YTS dataset is provided in a tabular format in CSV or Excel format and can be directly imported into R. The dataset was published on June 3, 2014, with the latest update on June 7, 2018. The metadata was last updated on August 25, 2023. It contains 10600 observations with 31 variables. We primarily focus on variables representing state, year, gender, education level, smoking status, type of tobacco used, and cessation efforts.

We also identified some potential issues with the data. First, there are some columns (TopicId, MeasuredId, StratificationID...) generated from the sampling process, which are irrelevant to the tobacco usage and cessation. Thus, these columns will be excluded in the analysis. Secondly, the dataset both gives the sample size of the youth who share the same smoking status under a given condition and the percentage of it among the total population. Considering the substantial variability in sample sizes based on different populations in schools and states, our emphasis will be on percentage values rather than sample size. Thirdly, in the gender column, the responses are female, male, and overall. Sample size of female and male do not exactly add up to the overall sample size, as other gender types are included. However,only binary genders (female and male) will be considered for comparisons, given the small sample sizes for non-binary genders. Additionally, due to missing data for certain states in specific years, temporal analysis may need to be conducted state-wise

### **Research plan**

In this section, we will first provide a brief introduction to our dependent and independent variables. We will then discuss the target values of interest that our data enables us to explore, followed by an examination of patterns that can be explored using demographic independent variables.

Our primary dependent variable is "Data_Value," representing the percentage of students who share the same smoking status among the entire student population with similar features. In other words, holding other features constant in an observation (Year, Location, education, and gender), "Data_Value" provides the percentage of students with the given smoking status, including type of tobacco, user status, and frequency response.

Regarding independent variables, we have general demographic variables such as gender, education, states, and year of the survey. Additionally, we have variables related to smoking status, including the type of tobacco (smokeless cigarettes or not), user status (attempted/thought about quitting or not), and frequency response (how frequently one smokes).

For tobacco usage, we are interested in students' exposure to tobacco products and their frequency of usage. To explore middle school and high school students' exposure to tobacco and smokeless tobacco, respectively, we will focus on the subset of observations where the Response variable is equal to "Ever." According to the survey design, students who answered "Ever" reported having tried smoking cigarettes or smokeless cigarettes. We will also control for the TopicDesc variable to be either "Cigarette Use (Youth)" or "Smokeless Tobacco Use (Youth)" to examine student exposure to tobacco and smokeless tobacco separately. The corresponding values in the "Data_Value" variable can then be interpreted as the percentage of students who reported ever trying smoking cigarettes or smokeless cigarettes.

On the other hand, controlling for other variables, we can also explore the frequency of smoking among the youth population. Instead of focusing on those who answered "Ever" to the Response variable, we will concentrate on those who reported "Current" and "Frequent." The corresponding values in the "Data_Value" variable for those who answered "Current" represent the percentage of students who had smoked on one or more days in the 30 days prior to the survey. Those who answered "Frequent" reported having smoked on 20 or more of the 30 days preceding the survey. We can again use the TopicDesc variable to examine cigarette usage and smokeless cigarette usage separately.

Concerning tobacco cessation, we are also interested in understanding students' willingness to quit tobacco usage and their attempts. Although the dataset has a limited sample size, it provides relevant information. By controlling the TopicDesc variable to be "Cessation (Youth)" and the MeasureDesc variable to be "Quit Attempt in Past Year Among Every Day Cigarette Smokers," the observations in the "Data_Value" variable represent the percentage of current smokers who quit smoking for one day or more in the past year. Changing the MeasureDesc variable to "Percent of Current Smokers Who Want to Quit" allows the observation in the "Data_Value" variable to represent the percentage of current smokers who want to stop smoking cigarettes.

While using smoking status-related dependent variables to explore tobacco usage prevalence among the youth population and their desire to quit smoking, we will employ demographic independent variables such as States, Gender, Level of education, and year of the survey to investigate potential disparities in patterns of tobacco usage and cessation in the youth population. Specifically, we can:

1.  Investigate youth tobacco usage and cessation across states by changing the Location variable while controlling for other variables.

2.  For each given state and year of the survey, investigate potential differences in youth tobacco usage and cessation between genders using the Gender variable.

3.  Similarly, investigate the difference between the prevalence of tobacco usage and cessation in middle school and high school using the Education variable.

4.  Vary the Year variable to observe the evolution of youth tobacco usage and cessation from 1999 to 2017 in chosen states or across all states. Additionally, we can explore how the difference in gender patterns of tobacco usage and cessation changes across years.

### Missing Value Analysis

```{r}
# import dataset
library(readr)
data <- read_csv("https://raw.githubusercontent.com/yutongzhang37/tobacco/main/Data/Youth_Tobacco_Survey__YTS__Data.csv")
```

We will remove variables with single unique observations and variables that are repetitive in nature.

```{r,message=FALSE}
library(tidyverse)

# We now remove variables with single observation 
data <- Filter(function(x)(n_distinct(x)>1), data)

# use only the independent and dependent variable mentioned above
data_bf <- subset(data, select = c("YEAR", "LocationAbbr", "TopicDesc", "MeasureDesc", "Response","Data_Value","Sample_Size", "Gender", "Education"))

# Number of missing values in different columns
colSums(is.na(data_bf)) %>%
  sort(decreasing = TRUE)
```

Only three variables contain missing values: Response, Data_Value, and Sample_Size. We now look for missing value patterns.

```{r,message=FALSE}
library(mi)
x<- missing_data.frame(as.data.frame(data_bf))
image(x,greyscale= FALSE)
```

From above graph, we can clearly see that Response, Data_value, Sample_Size are the only three variables with missing values. Data_Value and Sample_Size seem to have the same missing value pattern. In other words, when Sample_Size is missing, Data_Value is also missing, which is reasonable to observe.

However, since we have a very large dataset, heapmap is the not the best to visualize the missing value patters. We will use plot_missing() function.

```{r,message=FALSE}
# | fig-width: 30
# | out-width: 100%
library(redav)

# shorten variable name
data_bf <- data_bf %>% 
  rename(
    State = LocationAbbr,
    Topic = TopicDesc, 
    Measure = MeasureDesc, 
    Sample = Sample_Size, 
    Edu = Education,
    Res = Response,
    Value = Data_Value
)
plot_missing(data_bf, percent = FALSE)
```

From the upper marginal graph, it is evident that the variable "Response" contains a larger number of missing values. This occurrence is attributed to one of the "TopicDesc" values being set to "Cessation," which aims to capture students' willingness and attempts to quit tobacco usage. Consequently, the "Response" variable, which is utilized to record the frequency of tobacco usage, is left as missing in the dataset.

From the main graph, we can confirm that Data_Value and Sample_Size share the same missing value pattern. In pattern 2, the sole missing value is observed in the "Response" variable, as mentioned earlier, due to its purpose in capturing tobacco cessation.

In pattern 3, all three variables---Response, Data_Value, and Sample_Size---are missing. Further investigation, referencing the code book and dataset, reveals that this situation arises from a very limited number of samples. This is predominantly the case when narrowing down the analysis to either females or males, coupled with the "TopicDesc" value set to "Cessation."

Pattern 4, where only "Data_Value" and "Sample_Size" are missing, does not exhibit such occurrences. Verification can be conducted by the following:

```{r}
data_bf |> filter(Value == NA & Sample == NA & Res != NA)
```
