# Results

In this section, we will first briefly introduce variables used for analysis and then discuss ...**... TBA**

### Data Processing

We will continue to use variables mentioned previously.

-   Dependent variable:

    -   Value : the percentage of students who share the same smoking status among the entire student population with similar features

-   Independent variable:

    -   YEAR: Year of the observation

    -   State: Location of the observation

    -   Topic: Type of tobacco consumption(Cigarette/ Smokeless Tobacco/ Cessation)

    -   Measure: Smoking Status (Smoking/ Quit Attemp)

    -   Res: Frequency of tobacco usage (Ever/ Current/ Frequent)

    -   Sample: sample size

    -   Edu: Education background of the observation (Middle School/ High School)

    -   Gender: Female/ Male/ Overall

```{r, message=FALSE}
# Impart Packages 
library(readr)
library(tidyverse)
library(ggplot2)
library(choroplethr)
library(dplyr)
library(plotly)
library(cowplot)
library(vcd)
library(choroplethrMaps)
# Import Dataset 
data <- read_csv("https://raw.githubusercontent.com/yutongzhang37/tobacco/main/Data/Youth_Tobacco_Survey__YTS__Data.csv")

# We now remove variables with single observation
data <- Filter(function(x)(n_distinct(x)>1), data)

# use only the independent and dependent variable mentioned previously
data_bf <- subset(data, select = c("YEAR", "LocationDesc", "TopicDesc", "MeasureDesc", "Response","Data_Value","Sample_Size", "Gender", "Education"))

# Rename variables
data_bf <- data_bf %>%
  rename(
    State = LocationDesc,
    Topic = TopicDesc,
    Measure = MeasureDesc,
    Sample = Sample_Size,
    Edu = Education,
    Res = Response,
    Value = Data_Value
  )
```

### Tabacco Exposure and Frequent Smoker Variance by State in 2017

#### Choropleth Map

```{r}
# We fixed on the data from 2017 (the latest year in the dataset)
# to explore whether there is a difference bewteen tobacco exposure, usage frequency and cessation
# between different state/ education level/ gender.

# exposure
data2017 <- subset(data_bf, YEAR == 2017)
data2017 <- data2017 |>
  mutate(State = tolower(State)) |>
  rename(region = State, value = Value)

exposure <- data2017 |>
  filter(Topic == "Cigarette Use (Youth)" & Res == "Ever") |>
  select(region, value, Gender, Edu)
exposure_middle <- exposure |>
  filter(Gender == "Overall" & Edu == "Middle School") |>
  select(region, value)
exposure_high <- exposure |>
  filter(Gender == "Overall" & Edu == "High School") |>
  select(region, value)
suppressWarnings({
  state_choropleth(exposure_middle, num_colors = 1, title = "2017 Middle School Youth Cigarette Exposure") +
    annotate("text", x = -104, y = 23, label = "HI", color = "black", size = 3) +
    annotate("text", x = -116, y = 23, label = "AK", color = "black", size = 3)
})
suppressWarnings({
  state_choropleth(exposure_high, num_colors = 1, title = "2017 High School Youth Cigarette Exposure") +
    annotate("text", x = -104, y = 23, label = "HI", color = "black", size = 3) +
    annotate("text", x = -116, y = 23, label = "AK", color = "black", size = 3)
})
```

```{r}
# frequent smoker
frequent <- data2017 |>
  filter(Topic == "Cigarette Use (Youth)" & Res == "Frequent") |>
  select(region, value, Gender, Edu)
frequent_middle <- frequent |>
  filter(Gender == "Overall" & Edu == "Middle School") |>
  select(region, value)
frequent_high <- frequent |>
  filter(Gender == "Overall" & Edu == "High School") |>
  select(region, value)
suppressWarnings({
  state_choropleth(frequent_middle, num_colors = 1, title = "2017 Middle School Frquent Smoker") +
    annotate("text", x = -104, y = 23, label = "HI", color = "black", size = 3) +
    annotate("text", x = -116, y = 23, label = "AK", color = "black", size = 3)
})
suppressWarnings({
  state_choropleth(frequent_high, num_colors = 1, title = "2017 High School Youth Frquent Smoker") +
    annotate("text", x = -104, y = 23, label = "HI", color = "black", size = 3) +
    annotate("text", x = -116, y = 23, label = "AK", color = "black", size = 3)
})
```

```{r}
# cessation
willingness <- data2017 |>
  filter(Topic == "Cessation (Youth)" & Measure == "Percent of Current Smokers Who Want to Quit") |>
  select(region, value, Gender, Edu)
willingness_middle <- willingness |>
  filter(Gender == "Overall" & Edu == "Middle School") |>
  select(region, value)
willingness_high <- willingness |>
  filter(Gender == "Overall" & Edu == "High School") |>
  select(region, value)
suppressWarnings({
  state_choropleth(willingness_middle, num_colors = 1, title = "2017 Middle School Quit Smoking Williness") +
    annotate("text", x = -104, y = 23, label = "HI", color = "black", size = 3) +
    annotate("text", x = -116, y = 23, label = "AK", color = "black", size = 3)
})
suppressWarnings({
  state_choropleth(willingness_high, num_colors = 1, title = "2017 High School Quit Smoking Williness") +
    annotate("text", x = -104, y = 23, label = "HI", color = "black", size = 3) +
    annotate("text", x = -116, y = 23, label = "AK", color = "black", size = 3)
})
```

#### Cleveland Dot Plot

```{r}
# exposure
exposure_g_middle <- exposure |>
  filter(Gender != "Overall" & Edu == "Middle School")
exposure_g_high <- exposure |>
  filter(Gender != "Overall" & Edu == "High School")
plot1 <- ggplot(exposure_g_middle, aes(x = value, y = fct_reorder2(region, Gender=='Male', value, .desc=FALSE), color = Gender)) +
  geom_point() +
  labs(title = "2017 Middle School Youth",
       x = "Percentage",
       y = "State") +
  theme_minimal()
plot2 <- ggplot(exposure_g_high, aes(x = value, y = fct_reorder2(region, Gender=='Male', value, .desc=FALSE), color = Gender)) +
  geom_point() +
  labs(title = "2017 High School Youth",
       x = "Percentage",
       y = "State") +
  theme_minimal()
plot_grid(plot1, plot2, ncol = 1, align = "vh")
```

```{r}
# frequent smoker
frequent_g_middle <- frequent |>
  filter(Gender != "Overall" & Edu == "Middle School")
# fixing our target to frequent smokers, there is no data for binary gender in middle school
frequent_g_high <- exposure |>
  filter(Gender != "Overall" & Edu == "High School")
ggplot(frequent_g_high, aes(x = value, y = fct_reorder2(region, Gender=='Male', value, .desc=FALSE), color = Gender)) +
  geom_point() +
  labs(title = "2017 High School Frquent Smoker by Gender",
       x = "Frequent Smoker Percentage",
       y = "State") +
  theme_minimal()
```

#### Mosaic Plot for Louisiana in 2017

```{r}

# Exposure frequency
exposure_mosaic <- data2017 |>
  filter(region == "louisiana" & Topic == "Cigarette Use (Youth)" & Res == "Ever" & Gender != "Overall")|>
  mutate(Exposure = ifelse(Res == "Ever", "Had Smoked", Res)) |>
  select(Gender, Edu, Exposure, value, Sample) |>
  mutate(Frequency = round(0.01 * value * Sample))
never_smoked_mosaic <- exposure_mosaic |>
  mutate(Exposure = "Never Smoked",
         Frequency = Sample - Frequency)
exposure_mosaic_full <- bind_rows(exposure_mosaic, never_smoked_mosaic) |>
  select(-value, -Sample) |>
  rename(Education_Level = Edu)
exposure_mosaic_full$Education_Level <- factor(exposure_mosaic_full$Education_Level, levels = c("Middle School", "High School"))
exposure_mosaic_full$Exposure <- factor(exposure_mosaic_full$Exposure, levels = c("Never Smoked", "Had Smoked"))
exposure_mosaic_cases <-tidyr::uncount(base::as.data.frame(exposure_mosaic_full), Frequency)
vcd::mosaic(Exposure ~ Education_Level + Gender, data = exposure_mosaic_cases, direction = c("v","v","h"),
            highlighting_fill = c("lightyellow", "lightblue"), gp_labels=(gpar(fontsize=9)))
```

```{r}
# frequent smoker percentage
frequent_mosaic <- data2017 |>
  filter(region == "louisiana" & Topic == "Cigarette Use (Youth)" & Res == "Frequent" & Gender != "Overall")|>
  mutate(Frequent_Smoker = ifelse(Res == "Frequent", "Yes", Res)) |>
  select(Gender, Edu, Frequent_Smoker, value, Sample) |>
  mutate(Frequency = round(0.01 * value * Sample))
not_frequent_mosaic <- frequent_mosaic |>
  mutate(Frequent_Smoker = "No",
         Frequency = Sample - Frequency)
frequent_mosaic_full <- bind_rows(frequent_mosaic, not_frequent_mosaic) |>
  select(-value, -Sample) |>
  rename(Education_Level = Edu)
frequent_mosaic_full$Education_Level <- factor(frequent_mosaic_full$Education_Level, levels = c("Middle School", "High School"))
frequent_mosaic_full$Frequent_Smoker <- factor(frequent_mosaic_full$Frequent_Smoker, levels = c("No", "Yes"))
frequent_mosaic_cases <-tidyr::uncount(base::as.data.frame(frequent_mosaic_full), Frequency)
vcd::mosaic(Frequent_Smoker ~ Education_Level + Gender, data = frequent_mosaic_cases, direction = c("v","v","h"),
            highlighting_fill = c("lightyellow", "lightblue"), gp_labels=(gpar(fontsize=9)))
```
